Estimate Response to Selection

### 2021-07-01 Manuscript/manuscript.Rmd
update methods and response to selection results

One of the fundamental questions in Ecology is To what extent does biodiversity
drive variation in ecosystem functioning? Variation in abiotic conditions
regulates the rate of ecosystem processes, but it remains unclear to what
extent variation in community structure regulates the rate of ecosystem
processes. Many ecosystem functions are mediated by organisms. In particular,
microorganisms such as bacteria mediate the major biogeochemical cycles such
as the carbon and nitrogen cycles. Despite the central role of bacteria in
performing these processes, we have not found a consistent relationship between
microbial community structure and the rate of any ecosystem process. In order
to understand the importance of biodiversity to ecosystem functioning and to
better predict and manage specific ecosystem processes, we need to determine
the relationship between microbial community structure and ecosystem function.

In macroorganismal communities, there appears to be a relationship between the
diversity of a community and the resilience of ecosystem functioning to
environmental perturbations (Tilman, ????... Hooper). Typically, this is the
diversity (or more specifically, the richness) of a particular functional
group, such as photosynthetic plants. Similar studies in microbial communities
rarely find a relationship between diversity and function (Graham, ). However,
these studies typically define diversity much more broadly than plant diversity
studies by looking at the phylogenetic diversity of the entire bacterial and
archaeal community, not just of a particular functional group. Studies that
focus more narrowly on a particular ecosystem process and a functional group
involved in that process still rarely find a significant correlation between
the abundance of a functional group (or a functional protein-coding gene) and
the rate of the corresponding process (Rocca, ????). For certain ecosystem
functions in certain ecosystems, these relationships can be informative. For
example, the abundance of a single methanogenic species (*G. species*) is
correlated with methane emissions in a permafrost ecosystem (cite). In addition,
bacterial and archaeal community composition is correlated with methane
emissions in forests and pastures of the Brazilian Amazon (Meyer et al. 2020).
Therefore, there appears to be some indication that microbial community
structure drives variation in methane cycling in soils.

To further investigate the relationship between microbial community structure
and the rate of methane cycling in soils, we want to determine which organisms
are important for regulating methane cycling. This ecological question is
analogous to the problem of genotype-phenotype mapping in quantitative genetics
(Morris 2020). Variation in an organimsal trait is partially determined by
environmental conditions and partially determined by genetic varition. One goal
of quantitative genetics is to determine the proportion of the variation
attributable to genetics and to identify which alleles regulate the phenotypic
trait. These goals are commonly achieved through artificial selection
experiments and association mapping studies, such as genome-wide association
studies. To determine the mapping between microbial community structure and
ecosystem function, we applied these approaches of genotype-phenotype mapping
to whole microbial communities. In this case, the "alleles" are different
microbial taxa and the "phenotype" is the rate of ecosystem function at the
whole-ecosystem level.

To do this, we performed an artificial ecosystem selection experiment to
identify communities of bacteria and archaea that collectively perform a high
rate of methane oxidation (Swenson et al. 2000).  (Goodnight et al. 1997,
Williams and Lenton 2007)we generated twenty-four soil ecosystems in the lab
using a sterilized potting mix as the substrate. We then inoculated these
ecosystems with a small amount of living soil to generate variation in
community structure across the twenty-four ecosystems. These ecosystems were
then assigned to one of two selection treatments: neutral selection with
ecosystems chosen for reproduction at random or positive selection with
ecosystems chosen for reproduction based on their methane oxidation rate. Each
ecosystem was maintained at 1000 ppm CH4 over several weeks to allow for
colonization and growth of the microbial community and to enrich for methane
oxidizers. We then determined the methane oxidation rate for each ecosystem and
selected three jars in each treatment to inoculate the next set of twenty-four
jars. This process continued for five passages. At the end of the experiment,
we extracted DNA from soils in the second and fifth passage for both the
neutral and positive treatments. By comparing the response to selection within
the community between the two treatments, we can identify microbial species
that contribute to variation in ecosystem methane emissions.

This approach is powerful because by using a common soil substrate and
maintaining a constant headspace concentration of methane, we can eliminate
much of the environmental variation that would be present in an observational
study in the wild. In addition, by applying selection on methane oxidation rate
in the positive treatment, we can enrich for taxa involved in methane cycling
making it easier to perform association mapping between microbial taxa and
ecosystem methane emissions. In this paper, we address whether there is a
response to selection on methane oxidation rate at the whole ecosystem level. We
then quantify the amount of variation in methane oxidation attributable to
variation in community composition within our laboratory environment. Finally,
we identify microbial markers of soil methane cycling using association mapping
in order to deduce the mapping between microbial community structure and
ecosystem function.

@@

## Selection experiment
We performed an artificial ecosystem selection experiment similar to Swenson et
al. (2000) by passaging replicate soil microbial communities. The trait we
selected on was CH~4~ oxidation rate. Our experiment had two selection lines
with twelve jars each for a total of twenty-four jars per passage. One line was
a positive selection line where the two or three jars with the highest CH~4~
oxidation rate were chosen to inoculate the next set of jars. The other line
was a neutral selection line where an equal number of jars as the positive line
were chosen at random among the twelve. The number of jars chosen was based on
the distribution of fluxes among the positive jars, i.e., we chose the top
three jars unless only two jars had considerably greater oxidation rates based
on visual inspection of histograms.

The initial soil microbial community was sampled from the top 10 cm of an
upland mineral soil under a deciduous forest ecosystem near the University of
Oregon campus. Incubations were performed in 500 mL mason jars (Ball
Corporation, Broomfield, Colorado, USA) with rubber septa installed in the
lids. Jars were incubated at ambient temperature on the benchtop. To create the
first set of jars, twenty-four jars were created in an identical manner and
then randomly assigned to either the positive or negative selection line. Each
jar was sterilized with 70% ethanol to which was added 45 g of autoclaved
potting mix, 5 g of live soil, and 3.5 mL of sterile deionized water to bring
them to <mark> 60%??? of field capacity </mark>. The jars were then
homogenized, capped, and injected with 4.3 mL of 99% CH~4~ to bring the
headspace concentration to approximately 1000 ppm CH~4~. Jars were flushed and
respiked twice per week to maintain elevated CH~4~ concentrations until travel,
holidays, and COVID-19 got in the way and then we just did it as often as we
could. Jars were incubated for approximately three weeks and then flux
measurements were taken.

Once CH~4~ oxidation rates were determined, the top two or three jars were
chosen for selection in the positive line and an equal number of jars were
randomly chosen from the neutral line. The selected jars for each line were
homogenized and used as the 5 g live soil for the subsequent passage. Twelve
new jars were created for each line in the same manner as above using the
inoculum for that line. This procedure continued until a divergence was
observed in the mean flux of each line for two consecutive generations. Jars in
passage number 2 and 5 were homogenized and a sample was collected and stored
at -80$^\circ$C for later DNA extraction.

## Methane flux measurements Methane oxidation rates were determined after
flushing and spiking jars to approximately 1000 ppm CH~4~. Head-space samples
of 1 mL were collected from each jar immediately after spiking and then at time
points 3, 6, 24, and 48 hours for a 5-point curve. Samples were immediately
injected into a <mark>???</mark> gas chromatograph fitted with an electron
capture detector (company, city, state, USA) to determine the headspace CH~4~
concentration. Fluxes were calculated from a first-order exponential decay
function as decay constant `k`. Oxidation rates are presented as the additive
inverse of `k` (i.e., `-k`) so that a more positive value represents a faster
oxidation rate.

@@

Soil DNA was extracted from a 0.25 g subsample of soil from the starting
inoculum and from each jar in passage 2 and 5. Negative controls were extracted
from autoclaved potting mix and DNase-free water. Extractions were performed
using the DNeasy PowerSoil kit (QIAGEN, Hilden, Germany) and quantified using
Qubit (company, city, state). In order to estimate the diversity and relative
abundance of the bacterial and archaeal taxa in our soil ecosystems, we
sequenced the V3-V4 region of the 16S rRNA gene using the Earth Microbiome
Project protocol and primers<mark>???</mark>. In addition, to estimate the
diversity and relative abundance of CH~4~ oxidizing taxa, we sequenced the
particulate methane monooxygenase subunit A gene (*pmoA*) using
primers<mark>???</mark>. Sequencing was performed on the Illumina MiSeq in v3
mode with 300 bp paired-end reads.

## Bioinformatics Bioinformatic analyses were conducted using the `QIIME2`
bioinformatics platform (Bolyen et al. 2019). Raw sequencing reads were
demultiplexed using ??? (citation), denoised using `DADA2` (citation).
Taxonomic assignment for the 16S reads was performed using ??? based on the
Silva database. All subsequent tatistical analyses were performed in the `R`
statistical programming environment (R Core Team 2020). Richness estimates were
made using the `breakaway` package and diversity indices using the `DivNet`
package (Willis et al. ???).

@@

To test whether there was a significant change in CH~4~ oxidation rate (k) as a
response to selection, we tested the deviation between the positive and neutral
selection lines over multiple passages. We calculated deviation as the
difference in the mean oxidation rate between the positive and neutral
treatment within each passage. The standard errors of the difference were
calculated as the square root of the sum of the squared errors of the positive
and neutral lines. We then fit an ordinary least squares regression model
taking into account the standard errors of the deviations. The model was fit
using the `rma` function from the `metafor` package in `r` (Viechtbauer 2010).

@@

The intercept for the Neutral treatment was significantly different from zero
at `r myround(tidy(response_model)[[1, 2]], 2)` (SE = `r
myround(tidy(response_model)[[1, 3]], 2)`, t = `r myround(tidy(response_model)[[1,
4]], 2)`, p = `r myround(tidy(response_model)[[1, 5]], 2)`). The Positive
treatment started at a significantly lower methane oxidation rate than the
Neutral treatment with a y-intercept of `r myround(tidy(response_model)[[3, 2]],
2)` relative to Neutral (SE = `r myround(tidy(response_model)[[3, 3]], 2)`, t = `r
myround(tidy(response_model)[[3, 4]], 2)`, p = `r myround(tidy(response_model)[[3,
5]], 2)`). There was no change in methane oxidation rate of the Neutral
treatment over the five passages (slope = `r myround(tidy(response_model)[[2, 2]],
2)`, SE = `r myround(tidy(response_model)[[2, 3]], 2)`, t = `r
myround(tidy(response_model)[[2, 4]], 2)`, p = `r myround(tidy(response_model)[[2,
5]], 2)`). By contrast, there was a significant increase in the Positive
treatment of `r myround(tidy(response_model)[[4, 2]], 2)` per passage which is
approximately a `r round((pos_slope - 1) * 100, 0)`% increase in methane
oxidation rate per passage (SE = `r myround(tidy(response_model)[[4, 3]], 2)`, t =
`r myround(tidy(response_model)[[4, 4]], 2)`, p = `r myround(tidy(response_model)[[4,
5]], 2)`).

Methane oxidation rate increased as a response to selection in the Positive
selection treatment (`r figures("deviance", display="cite")`).  This can be
seen in the increase in deviation between the Positive and Neutral selection
lines over the course of the experiment (z = , p = ).  Mean CH~4~ oxidation
rate for the Positive treatment increased by  $\pm$ per passage relative to the
Neutral treatment. The mean flux across all jars in all treatments in the first
passage was  $\pm$  so this represents a doubling of CH~4~ consumption in every passage due
to selection.

>>>>

One of the fundamental questions in Ecology is To what extent does biodiversity
drive variation in ecosystem function? Variation in abiotic conditions
regulates the rate of ecosystem processes, but it remains unclear to what
extent variation in community structure regulates the rate of ecosystem
processes. Many ecosystem functions are mediated by organisms. In particular,
microorganisms such as bacteria mediate the major biogeochemical cycles such
as the carbon and nitrogen cycles. Despite the central role of bacteria in
performing these processes, we have not found a consistent relationship between
microbial community structure and the rate of an ecosystem process. In order
to understand the importance of biodiversity to ecosystem functioning and to
better predict and manage specific ecosystem processes, we need to determine
the relationship between microbial community structure and ecosystem function.

In macroorganismal communities, there appears to be a relationship between the
diversity of a community and the resilience of ecosystem functions to
environmental perturbations (Tilman, ????... Hooper). Typically, this is the
diversity (or more specifically, the richness) of a particular functional
group, such as photosynthetic plants. Similar studies in microbial communities
rarely find a relationship between diversity and function (Graham, ). However,
these studies typically define diversity much more broadly than plant diversity
studies by looking at the phylogenetic diversity of the entire bacterial and
archaeal community. Studies that focus more narrowly on a particular ecosystem
process and a functional group involved in that process still rarely find a
significant correlation between the abundance of a functional group and the
rate of the corresponding process (Rocca, ????). For certain ecosystem
functions in certain ecosystems, these relationships can be informative. For
example, the abundance of a single methanogenic species (*G. species*) is
correlated with methane emissions in a permafrost ecosystem (cite). In
addition, bacterial and archaeal community composition is correlated with
methane emissions in forests and pastures of the Brazilian Amazon (Meyer et al.
2020).  Therefore, there appears to be some indication that microbial community
structure drives variation in methane cycling in soils.

To further investigate the relationship between microbial community structure
and the rate of methane cycling in soils, we want to determine which organisms
are important for regulating methane cycling. This ecological question is
analogous to the problem of genotype-phenotype mapping in quantitative genetics
(Morris 2020). Variation in an organismal trait is partially determined by
environmental conditions and partially determined by genetic variation. One goal
of quantitative genetics is to determine the proportion of the variation
attributable to genetics and to identify which alleles regulate the phenotypic
trait. These goals are commonly achieved through artificial selection
experiments and association mapping studies, such as genome-wide association
studies. To determine the mapping between microbial community structure and
ecosystem function, we applied these approaches of genotype-phenotype mapping
to whole microbial communities. In this case, the "alleles" are different
microbial taxa and the "phenotype" is the rate of ecosystem function at the
whole-ecosystem level.

To identify communities of bacteria and archaea that collectively perform a
high rate of methane oxidation, we performed an artificial ecosystem selection
experiment  (Swenson et al. 2000).  We generated twenty-four soil ecosystems in the lab using a sterilized
potting mix as the substrate. We then inoculated these ecosystems with a small
amount of living soil to generate variation in community structure across the
twenty-four ecosystems. These ecosystems were then assigned to one of two
selection treatments: neutral selection or positive selection. Each ecosystem
was maintained at 1000 ppm CH~4~ over several weeks to allow for colonization and
growth of the microbial community and to enrich for methane oxidizers. We then
determined the methane oxidation rate for each ecosystem. To impose selection,
we selected three jars from the positive treatment with the highest rate of
methane oxidation, homogenized them, and used that soil to inoculate the next
set of jars. For the neutral treatment, we selected three jars at random to
inoculate the next set of jars. This process continued for five passages. At
the end of the experiment, we extracted DNA from soils in the second and fifth
passage for both the neutral and positive treatments. By comparing the response
to selection within the community between the two treatments, we can identify
microbial species that contribute to variation in ecosystem methane emissions.

This approach is powerful because by using a common soil substrate and
maintaining a constant headspace concentration of methane, we can eliminate
much of the environmental variation that would be present in an observational
study. In addition, by applying selection on methane oxidation rate
in the positive treatment, we can enrich for taxa involved in methane cycling
making it easier to perform association mapping between microbial taxa and
ecosystem methane emissions. In this paper, we address whether there is a
response to selection on methane oxidation rate at the whole ecosystem level. We
then quantify the amount of variation in methane oxidation attributable to
variation in community composition within our laboratory environment. Finally,
we identify microbial markers of soil methane cycling using association mapping
in order to deduce the mapping between microbial community structure and
ecosystem function.

(Goodnight et al. 1997, Williams and Lenton
2007)

@@

## Selection experiment
We performed an artificial ecosystem selection experiment similar to Swenson et
al. (2000) by passaging replicate soil microbial communities. The trait we
selected on was CH~4~ oxidation rate. Our experiment had two selection lines
with twelve jars each for a total of twenty-four jars per passage. One line was
a positive selection line where the two or three jars with the highest CH~4~
oxidation rate were chosen to inoculate the next set of positive jars. The
other line was a neutral selection line where an equal number of jars as the
positive line were chosen at random to inoculate the next set of neutral jars.
The number of jars chosen was based on the distribution of fluxes among the
positive jars, i.e., we chose the top three jars unless only two jars had
considerably greater oxidation rates based on visual inspection of histograms.
The experiment was carried out over five passages.

The initial soil microbial community was sampled from the top 10 cm of an
upland mineral soil under a deciduous forest ecosystem near the University of
Oregon campus in Eugene, OR, USA. Incubations were performed in 500 mL mason
jars (Ball Corporation, Broomfield, Colorado, USA) with rubber septa installed
in the lids. Each jar was sterilized with 70% ethanol to which was added 45 g
of autoclaved potting mix, 5 g of living soil, and 3.5 mL of sterile deionized
water to bring them to <mark> 60%??? of field capacity</mark>. The jars were
then homogenized, capped, and injected with 4.3 mL of 99% CH~4~ to bring the
headspace concentration to approximately 1000 ppm CH~4~. To create the two
treatment groups, twenty-four jars were created in an identical manner and then
randomly assigned to either the positive or neutral selection line. Jars were
flushed and respiked twice per week to maintain elevated CH~4~ concentrations
and were incubated at ambient temperature for approximately three weeks.
Methane oxidation rates were determined at the end of the incubation period.
These values were then used to determine the jars selected to inoculate the
next set of jars.

## Methane oxidation rate measurements

Methane oxidation rates were determined after flushing and spiking jars to
approximately 1000 ppm CH~4~. Headspace samples of 1 mL were collected from
each jar immediately after spiking and then at time points 3, 6, 24, and 48
hours for a 5-point curve. Samples were immediately injected into a
<mark>???</mark> gas chromatograph fitted with an electron capture detector
(company, city, state, USA) to determine the headspace CH~4~ concentration.
Fluxes were calculated from a first-order exponential decay function as decay
constant `k` with units day^-1^. Oxidation rates are presented as the additive
inverse of `k` (i.e., `-k`) so that a more positive value represents a faster
oxidation rate. The jars selected for the positive treatment in passage 2 had
the lowest methane oxidation rate of the twelve jars due to a calculation error
in the methane oxidation rate. All other passages correctly used the top three jars.

@@

A subsample of soil from the starting inoculum and from every jar in passages 2
and 5 was collected and stored at -80$^\circ$C for later DNA extraction.
Soil DNA was extracted from a 0.25 g subsample of soil. Negative controls were extracted
from autoclaved potting mix and DNase-free water. Extractions were performed
using the DNeasy PowerSoil kit (QIAGEN, Hilden, Germany) and quantified using
Qubit (company, city, state). In order to estimate the diversity and relative
abundance of the bacterial and archaeal taxa in our soil ecosystems, we
sequenced the V3-V4 region of the 16S rRNA gene using the Earth Microbiome
Project protocol and primers<mark>???</mark>. Sequencing was performed on the
Illumina NovaSeq 6000 with paired-end 150 bp reads.

## Bioinformatics

Bioinformatic analyses were conducted using the `QIIME2`
bioinformatics platform (Bolyen et al. 2019). Raw sequencing reads were
demultiplexed using ??? (citation), denoised using `DADA2` (citation).
Taxonomic assignment for the 16S reads was performed using ??? based on the
Silva database. All subsequent tatistical analyses were performed in the `R`
statistical programming environment (R Core Team 2020). Richness estimates were
made using the `breakaway` package and diversity indices using the `DivNet`
package (Willis et al. ???).

@@

To test whether there was a significant change in CH~4~ oxidation rate (k) as a
response to selection, we tested a difference in slopes between the positive
and neutral selection lines. The CH~4~ oxidation rates were strongly right
skewed with most values close to zero and few large, positive values. This
resulted in residuals that did not meet the assumptions of constant variance
and normal distribution. Therefore, methane oxidation rates were log~10~
transformed to better meet the assumptions of a linear model and to make
figures easier to interpret. First, we tested the effect of treatment by
fitting two nested models with and without treatment using `lm` in `R`. The
model formulation for the full model was `flux ~ passage * treatment` and the
model formulation for the reduced model was `flux ~ passage`. We compared these
models using the likelihood ratio test with the `anova` function in `R`. We
then fit the full model to determine the slope of the positive line, which
represents the change in methane oxidation rate per passage as a response to
selection.

@@

We observed a response to selection on soil methane oxidation rate at the whole
ecosystem level (`r figures("response", display="cite")`).
At the start of the experiment, the Positive
treatment had a mean methane oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`%
lower than the
Neutral treatment (SE = 
`r myround(tidy(response_model)[[3, 3]], 2)`, t = 
`r myround(tidy(response_model)[[3, 4]], 2)`, p = 
`r myround(tidy(response_model)[[3, 5]], 2)`). There was no change in methane oxidation rate of the Neutral
treatment over the five passages (slope = 
`r myround(tidy(response_model)[[2, 2]], 2)`, SE = 
`r myround(tidy(response_model)[[2, 3]], 2)`, t = 
`r myround(tidy(response_model)[[2, 4]], 2)`, p = 
`r myround(tidy(response_model)[[2, 5]], 2)`). By contrast, the Positive treatment had a
`r round(per_slope)`% increase in methane
oxidation rate per passage (slope = 
`r myround(tidy(response_model)[[4, 2]], 2)`, SE = 
`r myround(tidy(response_model)[[4, 3]], 2)`, t =
`r myround(tidy(response_model)[[4, 4]], 2)`, p = 
`r myround(tidy(response_model)[[4, 5]], 2)`).

### 2021-07-01 R/03-response.Rmd
update methods and response to selection results

```{r response_model}
# Fit full interaction model
response_model <- lm(log_ch4 ~ passage * treat, data = fluxes)

# No treatment model
no_treat_model <- lm(log_ch4 ~ passage, data = fluxes)

# Back calculate the slope of the positive treatment
pos_slope <- 10^response_model$coef["passage:treatp"]

@@

@@

```{r save_model}
save(response_model, pos_slope, file = '../Output/response_model.Rdata')
```

>>>>

```{r raw_model}
raw_model <- lm(ch4 ~ passage * treat, data = fluxes)

ggplot(raw_model, aes(x = .fitted, y = .resid)) + geom_point()

@@

```{r response_model}
response_model <- lm(log_ch4 ~ passage * treat, data = fluxes)

ggplot(response_model, aes(x = .fitted, y = .resid)) + geom_point()

```{r no_treat_model}
# No treatment model
no_treat_model <- lm(log_ch4 ~ passage, data = fluxes)

# Back calculate the slope of the positive treatment
pos_slope <- 10^response_model$coef["passage:treatp"]
per_slope <- (10^response_model$coef["passage:treatp"] - 1) * 100
```

@@

```{r save_model}
save(response_model, per_slope, file = '../Output/response_model.Rdata')
```
